Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-02 src/tex/ms.tex
First edits to 2.1

\section{Method} \label{sec:method}

@@

In order to recalibrate the model, we have to understand what parameters the model has.
Here we give a succinct summary of the IMRPhenomD model and the relevant parameters.
For interested readers, please refer to \citep{khan2016frequency} for more details on construction of the model.

The IMRPhenomD model is constructed by combining three individually fitted parts
into one coherent waveform model, which consists of the inspiral, intermediate, and merger-ringdown part, 
\begin{align}\label{eq:joint_waveform}
	h(f,\theta,\Lambda^i)=h_{\mathrm{ins}}(f,\theta,\Lambda^i) + h_{\mathrm{int}}(f,\theta,\Lambda^i) + h_{\mathrm{rd}}(f,\theta,\Lambda^i).
\end{align}
Instead of fitting the strain, which is a highly oscillatory function that is
difficult to fit, the amplitude and phase are fitted since they are smoother
functions. In each part, the amplitude and phase are made using simple functions
of frequency such as polynomials or lorentzians. Specifically, the merger-ringdown 
amplitude is fitted by a lorentzian and the other parts are fitted using polynomials. 
\begin{equation}\label{eq:amplitude}
\begin{aligned}
	A_0&\equiv\sqrt{\frac{2\eta}{3\pi^{1/3}}}f^{-7/6}\\
	A_{\mathrm{ins}}(f;\theta)&=A_{\mathrm{PN}}(f;\theta)+A_0\sum_{i=1}^3\rho_if^{(6+i)/3} \\
	A_{\mathrm{int}}(f;\theta)&=A_0(\delta_0+\delta_1f+\delta_2f^2+\delta_3f^3+\delta_4f^4) \\
	A_{\mathrm{rd}}(f;\theta)&=A_0\left[\gamma_1\frac{\gamma_3f_{\mathrm{damp}}}{(f-f_{\mathrm{RD}})^2+(\gamma_3f_{\mathrm{damp}})^2}e^{-\frac{\gamma_2(f-f_{\mathrm{RD}})}{\gamma_3f_{\mathrm{damp}}}}\right],  
\end{aligned}
\end{equation}
where $A_{\mathrm{PN}}$ is the post-newtonian expansion of the insprial amplitude up to order $A_0f^2$, $f_{\mathrm{damp}}$ is the damping frequency, and $f_{\mathrm{RD}}$ is the frequency at ringdown. The ansatzes for phase can be found in \citep{khan2016frequency}. These simple analytic functions
consists of parameters $\Lambda^i=\{\rho_i,\delta_i,\gamma_i\}$, which are defined as follows. 
\begin{align} \label{eq:Lambda}
	\Lambda^i&=\lambda_{00}^i+\lambda_{10}^i\eta \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)(\lambda_{01}^i+\lambda_{11}^i\eta+\lambda_{21}^i\eta^2) \nonumber \\ 
	&+(\chi_{\mathrm{PN}}-1)^2(\lambda_{02}^i+\lambda_{12}^i\eta+\lambda_{22}^i\eta^2) \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)^3(\lambda_{03}^i+\lambda_{13}^i\eta+\lambda_{23}^i\eta^2),
\end{align}
where $\lambda$ are fitting coefficients obtained during calibration, $\eta$ is
the symmetric mass ratio, and $\chi_{\mathrm{PN}}$ is the post-Newtonian spin
parameter, which is defined as 
\begin{align}
	\chi_{\mathrm{PN}}=\frac{m_1\chi_1+m_2\chi_2}{m_1+m_2}-\frac{38\eta}{113}(\chi_1+\chi_2).
\end{align}
Here, $m_{1,2}$ and $\chi_{1,2}$ are the primary and secondary mass and spin,
respectively. Finally, the individual segments are first connected directly using 
step functions. Then, by fixing coefficients in the intermediate segment, one can make 
the final waveform is continuous in its first derivative.

%\kw{Need to work on this paragraph}
Combining Eq.~\ref{eq:joint_waveform},~\ref{eq:amplitude}~and~\ref{eq:Lambda}, 
we can see that the entire waveform is non-linear in $\lambda$. A slightly 
inaccurate set of $\lambda$ can significantly affect the shape of the generated 
waveforms. Thus, having a set of accurate waveform coefficients is important 
and fundamental to having accurate GW models. Generally, waveform coefficients 
are obtained by calibrating with NR waveforms, which are waveforms computed 
using NR simulations. In the case of \citep{khan2016frequency}, they first 
obtain a set of $\Lambda$ by fitting model generated waveforms to Eq.~\ref{eq:amplitude} 
and the phase ansatzes. Repeating with different NR waveforms, they obtain multiple 
sets of $\Lambda$, and $\lambda$ are subsequently found by fitting against 
Eq. \ref{eq:Lambda}. Since the fitting procedure is done in a piece-wise manner, 
the correlations between different segments are omitted, which could limit the 
accuracy of the model. Also, since fitting was performed before connecting 
individual segments, the final waveform does not guarantee to achieve the optimal 
waveform. The connecting procedure alters the previously fitted waveform. 
Hence, the model generated waveforms contains additional inaccuracies.

Instead, we recalibrate coefficients jointly, which we can remove inaccuracies
and biases discussed above, and can improve model accuracy. In the past, due to
the complex nature of GW strains and piece-wise formalism of IMRPhenomD,
non-linear fitting was difficult to be performed in optimizing coefficients.
Hence, piece-wise optimization was done to obtain coefficients. However, with
{\ripple} and AD from \jax, gradients of IMRPhenomD can be easily obtained, thus
allowing the use of gradient-based algorithms for us to recalibrate the model.

>>>>

\section{Optimization Method} \label{sec:method}

In this section, we first briefly review the contruction of the IMRPhenomD model and discuss
how the calibration parameters enter the waveform.
We then outline the mismatch and how it can be used as a loss fucntion. 
Finally we discuss the gradient descent algorithm and our stopping criterion.

@@

% Paragraphs we need:
% 1. What is the model
% 2. How do the parameters enter the model
We start by giving a succinct summary of the IMRPhenomD model and the relevant parameters.
Interested readers should refer to \citep{khan2016frequency} for more details on construction of the model.

Aligned-spin, frequency domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda)}\,.
\end{align}
\te{Check minus sign}
Here $f$ is the frequnecy, $\theta$ are of intrinsic parameters of the binary, and $\Lambda$ is a set of additional
parameters which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
PhenomD uses the TaylorF2 model \citep{boyle2007high} which is known up to 3.5PN order.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the ampltude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stictched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

In practice the $\Lambda$ parameters are fit for each section independently i.e., intermediate coefficients are fit whilst ignoring the MR region.
Finally, to map the grid of tuned $\Lambda$ parameters back to the intrinsic parameter space, IMRhenomD uses the polynomial function:
\begin{align} \label{eq:Lambda}
	\Lambda^i&=\lambda_{00}^i+\lambda_{10}^i\eta \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)(\lambda_{01}^i+\lambda_{11}^i\eta+\lambda_{21}^i\eta^2) \nonumber \\ 
	&+(\chi_{\mathrm{PN}}-1)^2(\lambda_{02}^i+\lambda_{12}^i\eta+\lambda_{22}^i\eta^2) \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)^3(\lambda_{03}^i+\lambda_{13}^i\eta+\lambda_{23}^i\eta^2)\,,
\end{align}
where the $\lambda$'s are the fitting coefficients we are going to optimize below, $\eta$ is
the symmetric mass ratio, and $\chi_{\mathrm{PN}}$ is the post-Newtonian spin
parameter, which is defined as 
\begin{align}
	\chi_{\mathrm{PN}}=\frac{m_1\chi_1+m_2\chi_2}{m_1+m_2}-\frac{38\eta}{113}(\chi_1+\chi_2)\,.
\end{align}
Here, $m_{1,2}$ and $\chi_{1,2}$ are the primary and secondary mass and spin,
respectively.

% More precisely, $\Lambda$ by fitting model generated waveforms to Eq.~\ref{eq:amplitude} 
% and the phase ansatzes. Repeating with different NR waveforms, they obtain multiple 
% sets of $\Lambda$, and $\lambda$ are subsequently found by fitting against 
% Eq. \ref{eq:Lambda}. Since the fitting procedure is done in a piece-wise manner, 
% the correlations between different segments are omitted, which could limit the 
% accuracy of the model. Also, since fitting was performed before connecting 
% individual segments, the final waveform does not guarantee to achieve the optimal 
% waveform.
% Finally, the individual segments are first connected directly using 
% step functions. Then, by fixing coefficients in the intermediate segment, one can make 
% the final waveform is continuous in its first derivative.

% In order to recalibrate the model, we have to understand what parameters the model has.
% Here we give a succinct summary of the IMRPhenomD model and the relevant parameters.
% For interested readers, please refer to \citep{khan2016frequency} for more details on construction of the model.

% The IMRPhenomD model is constructed by combining three individually fitted parts
% into one coherent waveform model, which consists of the inspiral, intermediate, and merger-ringdown part, 
% \begin{align}\label{eq:joint_waveform}
% 	h(f,\theta,\Lambda^i)=h_{\mathrm{ins}}(f,\theta,\Lambda^i) + h_{\mathrm{int}}(f,\theta,\Lambda^i) + h_{\mathrm{rd}}(f,\theta,\Lambda^i).
% \end{align}
% Instead of fitting the strain, which is a highly oscillatory function that is
% difficult to fit, the amplitude and phase are fitted since they are smoother
% functions. In each part, the amplitude and phase are made using simple functions
% of frequency such as polynomials or lorentzians. Specifically, the merger-ringdown 
% amplitude is fitted by a lorentzian and the other parts are fitted using polynomials. 
% \begin{equation}\label{eq:amplitude}
% \begin{aligned}
% 	A_0&\equiv\sqrt{\frac{2\eta}{3\pi^{1/3}}}f^{-7/6}\\
% 	A_{\mathrm{ins}}(f;\theta)&=A_{\mathrm{PN}}(f;\theta)+A_0\sum_{i=1}^3\rho_if^{(6+i)/3} \\
% 	A_{\mathrm{int}}(f;\theta)&=A_0(\delta_0+\delta_1f+\delta_2f^2+\delta_3f^3+\delta_4f^4) \\
% 	A_{\mathrm{rd}}(f;\theta)&=A_0\left[\gamma_1\frac{\gamma_3f_{\mathrm{damp}}}{(f-f_{\mathrm{RD}})^2+(\gamma_3f_{\mathrm{damp}})^2}e^{-\frac{\gamma_2(f-f_{\mathrm{RD}})}{\gamma_3f_{\mathrm{damp}}}}\right],  
% \end{aligned}
% \end{equation}
% where $A_{\mathrm{PN}}$ is the post-newtonian expansion of the insprial amplitude up to order $A_0f^2$, $f_{\mathrm{damp}}$ is the damping frequency, and $f_{\mathrm{RD}}$ is the frequency at ringdown.

%\kw{Need to work on this paragraph}
Although initially independent, the stiching procedure means that each section
of the waveform intrinsically depends on the full set of $\lambda$s. 
A slightly inaccurate set of $\lambda$s can therefore lead to inaccuracies in
the generated waveforms. 
Thus, having a set of accurate waveform coefficients is crucial to the accuracy
of the GW models. 
Importantly, since the fitting was performed on the individual segments,
the final waveform is not guarenteed to have $\lambda$s close to global minima.

% Generally, waveform coefficients 
% are obtained by calibrating with NR waveforms, which are waveforms computed 
% using NR simulations. In the case of \citep{khan2016frequency}, they first 
% obtain a set of $\Lambda$ by fitting model generated waveforms to Eq.~\ref{eq:amplitude} 
% and the phase ansatzes. Repeating with different NR waveforms, they obtain multiple 
% sets of $\Lambda$, and $\lambda$ are subsequently found by fitting against 
% Eq. \ref{eq:Lambda}. Since the fitting procedure is done in a piece-wise manner, 
% the correlations between different segments are omitted, which could limit the 
% accuracy of the model. Also, since fitting was performed before connecting 
% individual segments, the final waveform does not guarantee to achieve the optimal 
% waveform. The connecting procedure alters the previously fitted waveform. 
% Hence, the model generated waveforms contains additional inaccuracies.

At the time of construction this piece-wise fitting was necessary since
$\lambda$ has 209 components making the optimization computationally prohibitive.
Here, we recalibrate the $\lambda$ coefficients jointly. 
This is made possible by the use of gradient-based optimization algorithms which are
enabled by AD from \jax and {\ripple}.
% , gradients of IMRPhenomD can be easily obtained, thus
% allowing the use of gradient-based algorithms for us to recalibrate the model.
